### 2016-05-12 text/Chapter05.tex
Context-aware systems edits (Chapter 3)

\paragraph{Semantics of sub-coeffecting.}
Although we do not include sub-coeffecting in the core flat coeffect calculus, it is an interesting
extension to consider. Semantically, sub-coeffecting drops some of the available contextual
capabilities (drops some of the implicit parameters or some of the past values). This can be
modelled by adding a (family of) lifting operation(s):
%
\begin{itemize}
 \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{itemize}
%
The axioms of flat coeffect algebra do not, in general, require that
$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
but the property holds for the three sample coeffect systems we consier. For systems with
the above property, the \ident{split} operation can be expressed in terms of lifting
(sub-coeffecting) as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),~
    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
\end{array}
\end{equation*}
%
The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object
mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map}
for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
are first and second projections from a two-element pair. To define the
$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
to throw away additional parts of the context and then transform the values in the
context.

This alternative definition is valid for our examples, but we do not use it for three reasons. First,
it requires making sub-coeffecting a part of the core definition. Second, this would be the only
place where our semantics uses a variable \emph{twice} (in this case $c$). Note therefore that
our use of an explicit \ident{split} means that the structure required by our semantics does not
need to provide variable duplication and our model could be embedded in linear or affine category.
Finally, explicit \ident{split} is similar to the definition that is needed for structural coeffects
in Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.

@@

\noindent
Without the indexing, the \ident{merge} operations implements \emph{zip} on Maybe values,
returning a value only when both values are present. The behaviour of the \ident{split}
operation is partly determined by the indices. When the input is \emph{dead}, both values have
to be dead (this is also the only solution of $\cclrd{\ident{D}} =\cclrd{r}\sqcap\cclrd{s}$), but when
the input is \emph{live}, the operation can perform implicit sub-coeffecting and drop one of
the values.

@@

\section{Conclusions}

>>>>

\paragraph{Semantics of subcoeffecting.}
Although we do not include subcoeffecting in the core flat coeffect calculus, it is an interesting
extension to consider. Semantically, subcoeffecting drops some of the available contextual
capabilities (drops some of the implicit parameters or some of the past values). This can be
modelled by adding a (family of) lifting operation(s):
%
\begin{itemize}
 \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{itemize}
%
The axioms of flat coeffect algebra do not, in general, require that
$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
but the property holds for the three sample coeffect systems we consier. For systems with
the above property, the \ident{split} operation can be expressed in terms of lifting
(subcoeffecting) as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),~
    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
\end{array}
\end{equation*}
%
The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object
mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map}
for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
are first and second projections from a two-element pair. To define the
$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
to throw away additional parts of the context and then transform the values in the
context.

This alternative definition is valid for our examples, but we do not use it for three reasons. First,
it requires making subcoeffecting a part of the core definition. Second, this would be the only
place where our semantics uses a variable \emph{twice} (in this case $c$). Note therefore that
our use of an explicit \ident{split} means that the structure required by our semantics does not
need to provide variable duplication and our model could be embedded in linear or affine category.
Finally, explicit \ident{split} is similar to the definition that is needed for structural coeffects
in Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.

@@

\noindent
Without the indexing, the \ident{merge} operations implements \emph{zip} on Maybe values,
returning a value only when both values are present. The behaviour of the \ident{split}
operation is partly determined by the indices. When the input is \emph{dead}, both values have
to be dead (this is also the only solution of $\cclrd{\ident{D}} =\cclrd{r}\sqcap\cclrd{s}$), but when
the input is \emph{live}, the operation can perform implicit subcoeffecting and drop one of
the values.

@@

\section{Summary}

### 2016-05-12 text/Chapter06.tex
Context-aware systems edits (Chapter 3)

The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the
exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.

@@

Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models
sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
it behaves as identity.

>>>>

The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the
exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
pointwise composition, as well as a relation $\cleq$ that defines subcoeffecting.

@@

Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models
subcoeffecting which can turn a context with a value into a dead context (second case); otherwise
it behaves as identity.

### 2016-05-12 text/Chapter08.tex
Context-aware systems edits (Chapter 3)

Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read
the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
information and conclusions split available contextual information.

@@

The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The
index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of
the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape
$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in
Section~\ref{sec:further-types}, we need to modify the domains and codomains of the \ident{merge},
\ident{split} and \ident{dup} operations as follows.
%
\begin{equation*}
\begin{array}{lcl}
 \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
 \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
 \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
\end{array}
\end{equation*}
%
The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
(in \ident{merge}, but also in \ident{dup} which models contraction).
This means that the splitting and merging the context can change the coeffect annotation:
%
\begin{equation*}
\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
\end{equation*}
%
For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
splitting and then merging a context may lose some of the available contextual capabilities
(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:further-rev-contr}
and Lemma~\ref{thm:further-rev-exch} relied on sub-coeffecting in the flat system.

@@

However, the operations $\ident{merge}$, $\ident{split}$ and $\ident{lift}$ can now all
use ordering. For example, the \ident{merge} operation previously returned
a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit sub-coeffecting
to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and
$\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
This means that we can compose the original \ident{merge} operation which returns greatest lower
bound with $\ident{lift}$ to get the (less precise) operation required here.

@@

The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some
transformation on one or two values in a context and then add the new value as a fresh variable
to the variable context. We omit sub-coeffecting, but it could be easily added following
the method used elsewhere.

>>>>

Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-understanding}, we can also read
the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
information and conclusions split available contextual information.

@@

The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The
index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of
the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape
$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in
Section~\ref{sec:further-types}, we need to modify the domains and codomains of the \ident{merge},
\ident{split} and \ident{dup} operations as follows.
%
\begin{equation*}
\begin{array}{lcl}
 \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
 \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
 \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
\end{array}
\end{equation*}
%
The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
(in \ident{merge}, but also in \ident{dup} which models contraction).
This means that the splitting and merging the context can change the coeffect annotation:
%
\begin{equation*}
\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
\end{equation*}
%
For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
splitting and then merging a context may lose some of the available contextual capabilities
(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:further-rev-contr}
and Lemma~\ref{thm:further-rev-exch} relied on subcoeffecting in the flat system.

@@

However, the operations $\ident{merge}$, $\ident{split}$ and $\ident{lift}$ can now all
use ordering. For example, the \ident{merge} operation previously returned
a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit subcoeffecting
to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and
$\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
This means that we can compose the original \ident{merge} operation which returns greatest lower
bound with $\ident{lift}$ to get the (less precise) operation required here.

@@

The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some
transformation on one or two values in a context and then add the new value as a fresh variable
to the variable context. We omit subcoeffecting, but it could be easily added following
the method used elsewhere.
